---
title: "Introduction to peRspective"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to peRspective}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

peRspective provides an R interface to the [FINOS Perspective](https://perspective.finos.org/) library, a high-performance WebAssembly-powered data visualization engine. It offers interactive pivot tables, cross-tabulations, and 12+ chart types that run entirely in the browser.

## Quick Start

```{r setup}
library(peRspective)
```

The simplest usage is to pass a data frame directly:

```{r basic, eval=FALSE}
perspective(mtcars)
```

This opens a fully interactive viewer with a settings panel where you can:

- Drag columns in/out of the view
- Switch between chart types (datagrid, bar, line, scatter, heatmap, etc.)
- Add group-by (row pivot) and split-by (column pivot) dimensions
- Create filters, sort, and change aggregations
- Write computed expressions

## Configuration

You can set an initial configuration programmatically:
```{r config, eval=FALSE}
perspective(mtcars,
  group_by = "cyl",
  columns = c("mpg", "hp", "wt"),
  plugin = "Y Bar",
  theme = "Pro Dark"
)
```

Users can still modify the view interactively even after initial configuration.

### Available Chart Types

- `"Datagrid"` — interactive data grid (default)
- `"Y Bar"` / `"X Bar"` — vertical / horizontal bar charts
- `"Y Line"` — line chart
- `"Y Area"` — area chart
- `"Y Scatter"` / `"XY Scatter"` — scatter plots
- `"Treemap"` — treemap
- `"Sunburst"` — sunburst chart
- `"Heatmap"` — heatmap

### Filtering and Sorting

```{r filter-sort, eval=FALSE}
perspective(iris,
  filter = list(c("Species", "==", "setosa")),
  sort = list(c("Sepal.Length", "desc"))
)
```

### Computed Expressions

```{r expressions, eval=FALSE}
perspective(mtcars,
  expressions = c('"hp" / "wt"'),
  columns = c("mpg", "hp", "wt", '"hp" / "wt"')
)
```

## Arrow IPC for Large Datasets

For datasets with 100k+ rows, use Arrow IPC serialization for better performance:

```{r arrow, eval=FALSE}
# Requires the arrow package
big_data <- data.frame(
  x = rnorm(100000),
  y = rnorm(100000),
  group = sample(letters, 100000, replace = TRUE)
)
perspective(big_data, use_arrow = TRUE)
```

## Shiny Integration

peRspective includes full Shiny support with a proxy interface for streaming updates:

```{r shiny, eval=FALSE}
library(shiny)

ui <- fluidPage(
  perspectiveOutput("viewer", height = "600px"),
  actionButton("add", "Add Data")
)

server <- function(input, output, session) {
  output$viewer <- renderPerspective({
    perspective(mtcars, plugin = "Y Bar", group_by = "cyl")
  })

  observeEvent(input$add, {
    proxy <- perspectiveProxy(session, "viewer")
    new_data <- mtcars[sample(nrow(mtcars), 5), ]
    psp_update(proxy, new_data)
  })

  # Capture user's interactive config changes
  observeEvent(input$viewer_config, {
    message("User changed config: ", str(input$viewer_config))
  })
}

shinyApp(ui, server)
```

### Proxy Functions

- `psp_update(proxy, data)` — append new rows
- `psp_replace(proxy, data)` — replace all data
- `psp_clear(proxy)` — clear all rows
- `psp_restore(proxy, config)` — apply a config
- `psp_reset(proxy)` — reset to defaults
